Covariate Analysis of MAYO Reprocessed counts Regressing out Diagnosis as well for TWAS

### 2019-09-18 code/Mayo_RegressDiagnosis_4_TWAS.Rmd
AWS CLI

---
title: Covariate and differential expression analysis of MAYO reprocessed counts
output: html_notebook
---
Date of analysis update: "`r date()`"

@@

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=FALSE}
##It is assumed your working directory is "~/ampad-DiffExp/gene_level_analysis"
## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(plyr)
library(tidyverse)
library(psych)
library(limma)
library(edgeR)
library(biomaRt)
library(RColorBrewer)
library(cqn)
library(glmnet)
library(synapser)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')
library(knitr)
library(doParallel)
library(foreach)
library(sva)
cl = makeCluster(detectCores()-2)
registerDoParallel(cl)
options(xtable.type="html")
# Source modified utility functions from limma and sva package
source('./utilityFunctions/parallelDuplicateCorrelation.R')
source('./utilityFunctions/irwsva.build.R')
source('./utilityFunctions/f.pvalue.R')
knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

@@

### Data preprocessing
```{r preprocess.data, echo=TRUE}
# Filter metadata
METADATA = as_tibble(METADATA, .name_repair = "unique") %>%
  dplyr::filter(ID %in% colnames(COUNT)) %>%
  dplyr::filter(!is.na(RIN)) %>%
  dplyr::filter(!is.na(PCT_INTRONIC_BASES)) %>%
  dplyr::filter(!is.na(AgeAtDeath)) %>%
  dplyr::filter(!is.na(Source)) %>%
  dplyr::filter(!(ID %in% SAMPLES.EXCLUDE)) %>%
  as.data.frame()
# Fix Tissue
METADATA$Tissue = gsub('CER', 'CBE', METADATA$Tissue)
# Fix AgeAtDeath
METADATA = METADATA %>%
  dplyr::mutate(AgeAtDeath = gsub("_or_above", "", AgeAtDeath))
# Fix Diagnosis
METADATA = METADATA %>%
  dplyr::mutate(Source.Diagnosis = Diagnosis) %>%
  dplyr::mutate(Source.Diagnosis = factor(Source.Diagnosis,
                                          levels = c('AD', 'Control', 'Pathologic Aging', 'PSP'),
                                          labels = c('AD', 'CONTROL', 'PATH_AGE', 'PSP'))) %>%
  dplyr::mutate(Diagnosis = gsub("Pathologic Aging", "OTHER", Diagnosis),
                Diagnosis = gsub("PSP", "OTHER", Diagnosis),
                Diagnosis = factor(Diagnosis, 
                                   levels = c("AD", "Control", "OTHER"),
                                   labels = c('AD', 'CONTROL', 'OTHER'))) %>%
  dplyr::mutate(Tissue.Diagnosis = paste(Tissue, Diagnosis, sep = '.'),
                Tissue.SourceDiagnosis = paste(Tissue, Source.Diagnosis, sep = '.'))
# Fix Gender
METADATA = METADATA %>%
  dplyr::mutate(Sex = Gender,
                Sex = factor(Sex, levels = c('F','M'), labels = c('FEMALE','MALE')))
# Get RIN square
METADATA = METADATA %>%
  dplyr::mutate(RIN2 = RIN^2)
# Add apoe4 genotype (0, 1, 2)
METADATA$APOE4 = 0
METADATA$APOE4[METADATA$ApoE %in% c(24, 34)] = 1
METADATA$APOE4[METADATA$ApoE %in% c(44)] = 2
METADATA$Tissue.APOE4 = paste(METADATA$Tissue, METADATA$APOE4, sep = '.')
# Match covariates to expression data
indToRetain = intersect(METADATA$ID, colnames(COUNT))
indRemoved = setdiff(colnames(COUNT), METADATA$ID)
COUNT <- COUNT[, indToRetain]
rownames(METADATA) = METADATA$ID
METADATA = METADATA[indToRetain,]
```
`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project.

@@

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering}
primaryVariable <- c("Tissue.SourceDiagnosis", "Tissue.Diagnosis", "Tissue.APOE4")
#_# ADD DIAGNOSIS HERE TO REGRESS FOR TWAS
FactorCovariates <- c("Donor_ID", "Source", "Sex", "FLOWCELL", "Tissue.SourceDiagnosis", "Tissue.Diagnosis", "Tissue.APOE4", "Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$ID
# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=6}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left', padding  = unit(c(18,2,2,18), 'mm'))
```
### Explore metadata
```{r data.explore, fig.width = 12, fig.height = 8}
my.theme = theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = Tissue.Diagnosis, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme
# AgeAtDeath
p[[2]] = ggplot(COVARIATES, aes(x = Tissue.Diagnosis, y = as.numeric(AgeAtDeath))) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeAtDeath') + my.theme
# PMI
p[[3]] = ggplot(COVARIATES, aes(x = Tissue.Diagnosis, y = as.numeric(PMI))) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI') + my.theme
# Ribosomal bases
p[[4]] = ggplot(COVARIATES, aes(x = Tissue.Diagnosis, y = as.numeric(PCT_RIBOSOMAL_BASES))) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Fraction of Ribosomal Bases') + my.theme
# Intronic bases
p[[5]] = ggplot(COVARIATES, aes(x = Tissue.Diagnosis, y = as.numeric(PCT_INTRONIC_BASES))) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Fraction Intronic Bases') + my.theme
# Intergenic bases
p[[6]] = ggplot(COVARIATES, aes(x = Tissue.Diagnosis, y = as.numeric(PCT_INTERGENIC_BASES))) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Fraction Intergenic Bases') + my.theme
multiplot(plotlist = p, cols = 3)
```

@@

### Sample clustering
PCA based clustering of samples
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$E.no.na, scale.=T, center = T)
# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])
plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Tissue.Diagnosis, c('Tissue', 'Diagnosis'), sep = '\\.')
p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
p <- p + theme_bw() + theme(legend.position="top") + facet_grid(Tissue+.~Diagnosis, scales = 'free_y')
p
```

Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
COVARIATES.tmp = (COVARIATES[,c(FactorCovariates), drop = F])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E.no.na)))
cols = WGCNA::labels2colors(COVARIATES.tmp);
WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

@@

### Sanity check
```{r residual.adj, fig.width=12, fig.height=8}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(expr, 
                                              COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
residualSigCovars[["PC_res"]][[2]]$plotData
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(expr))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
PCA of residual data
```{r decompse.normalise.data2.1, fig.height=10, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(expr, scale.=T, center = T)
# Plot first 4 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])
plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Tissue.Diagnosis, c('Region','Diagnosis'), sep = '\\.') %>% 
  dplyr::mutate(Region = factor(Region),
                Source = factor(Source))
p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Region~., scales = 'free_y')
p
```
Tree based clustering of residual data
```{r decompse.normalise.data2.2, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,FactorCovariates])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
tree = hclust(as.dist(t(expr)))
cols = WGCNA::labels2colors(COVARIATES.tmp);
WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

@@

```{r knitmd, eval=FALSE, cache=FALSE, include=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')
synLogin()
knit2synapse::knitfile2synapse(file = "~/TWAS/code/Mayo_RegressDiagnosis_4_TWAS.Rmd",
                                 owner = "syn20798012")
```

>>>>

---
title: Covariate Analysis of MAYO Reprocessed counts Regressing out Diagnosis as well for TWAS
author: 'JKG - Adapted from: Thanneer Perumal'
output: html_notebook
editor_options:
  chunk_output_type: console
---
Date of analysis update: "`r date()`"

@@

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=FALSE}
##It is assumed your working directory is "~/ampad-DiffExp/gene_level_analysis"
## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(plyr)
library(tidyverse)
library(psych)
library(limma)
library(edgeR)
library(biomaRt)
library(RColorBrewer)
library(cqn)
library(glmnet)
library(synapser)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')
library(knitr)
library(doParallel)
library(foreach)
library(sva)
cl = makeCluster(detectCores()-2)
registerDoParallel(cl)
options(xtable.type="html")
# Source modified utility functions from limma and sva package
source('../utilityFunctions/parallelDuplicateCorrelation.R')
source('../utilityFunctions/irwsva.build.R')
source('../utilityFunctions/f.pvalue.R')
knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

@@

### Data preprocessing
```{r preprocess.data, echo=FALSE}
# Filter metadata
METADATA = as_tibble(METADATA, .name_repair = "unique") %>%
  dplyr::filter(ID %in% colnames(COUNT)) %>%
  dplyr::filter(!is.na(RIN)) %>%
  dplyr::filter(!is.na(PCT_INTRONIC_BASES)) %>%
  dplyr::filter(!is.na(AgeAtDeath)) %>%
  dplyr::filter(!is.na(Source)) %>%
  dplyr::filter(!(ID %in% SAMPLES.EXCLUDE)) %>%
  as.data.frame()
# Fix Tissue
METADATA$Tissue = gsub('CER', 'CBE', METADATA$Tissue)
# Fix AgeAtDeath
METADATA = METADATA %>%
  dplyr::mutate(AgeAtDeath = gsub("_or_above", "", AgeAtDeath))
# Fix Diagnosis
METADATA = METADATA %>%
  dplyr::mutate(Source.Diagnosis = Diagnosis) %>%
  dplyr::mutate(Source.Diagnosis = factor(Source.Diagnosis,
                                          levels = c('AD', 'Control', 'Pathologic Aging', 'PSP'),
                                          labels = c('AD', 'CONTROL', 'PATH_AGE', 'PSP'))) %>%
  dplyr::mutate(Diagnosis = gsub("Pathologic Aging", "OTHER", Diagnosis),
                Diagnosis = gsub("PSP", "OTHER", Diagnosis),
                Diagnosis = factor(Diagnosis, 
                                   levels = c("AD", "Control", "OTHER"),
                                   labels = c('AD', 'CONTROL', 'OTHER'))) %>%
  dplyr::mutate(Tissue.Diagnosis = paste(Tissue, Diagnosis, sep = '.'),
                Tissue.SourceDiagnosis = paste(Tissue, Source.Diagnosis, sep = '.'))
# Fix Gender
METADATA = METADATA %>%
  dplyr::mutate(Sex = Gender,
                Sex = factor(Sex, levels = c('F','M'), labels = c('FEMALE','MALE')))
# Get RIN square
METADATA = METADATA %>%
  dplyr::mutate(RIN2 = RIN^2)
# Add apoe4 genotype (0, 1, 2)
METADATA$APOE4 = 0
METADATA$APOE4[METADATA$ApoE %in% c(24, 34)] = 1
METADATA$APOE4[METADATA$ApoE %in% c(44)] = 2
METADATA$Tissue.APOE4 = paste(METADATA$Tissue, METADATA$APOE4, sep = '.')
# Match covariates to expression data
indToRetain = intersect(METADATA$ID, colnames(COUNT))
indRemoved = setdiff(colnames(COUNT), METADATA$ID)
COUNT <- COUNT[, indToRetain]
rownames(METADATA) = METADATA$ID
METADATA = METADATA[indToRetain,]
```
`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project.

@@

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering}
primaryVariable <- c("Tissue.SourceDiagnosis", "Tissue.Diagnosis", "Tissue.APOE4")
#_# ADD DIAGNOSIS HERE TO REGRESS FOR TWAS
FactorCovariates <- c("Donor_ID", "Source", "Sex", "FLOWCELL", "Tissue.SourceDiagnosis", "Tissue.Diagnosis", "Tissue.APOE4", "Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI", "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")
# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$ID
# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, cache=FALSE, fig.width=10, fig.height=6}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left', padding  = unit(c(18,2,2,18), 'mm'))
```
### Explore metadata
```{r data.explore, cache=FALSE, fig.width = 12, fig.height = 8}
my.theme = theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))
# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = Tissue.Diagnosis, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme
# AgeAtDeath
p[[2]] = ggplot(COVARIATES, aes(x = Tissue.Diagnosis, y = as.numeric(AgeAtDeath))) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeAtDeath') + my.theme
# PMI
p[[3]] = ggplot(COVARIATES, aes(x = Tissue.Diagnosis, y = as.numeric(PMI))) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI') + my.theme
# Ribosomal bases
p[[4]] = ggplot(COVARIATES, aes(x = Tissue.Diagnosis, y = as.numeric(PCT_RIBOSOMAL_BASES))) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Fraction of Ribosomal Bases') + my.theme
# Intronic bases
p[[5]] = ggplot(COVARIATES, aes(x = Tissue.Diagnosis, y = as.numeric(PCT_INTRONIC_BASES))) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Fraction Intronic Bases') + my.theme
# Intergenic bases
p[[6]] = ggplot(COVARIATES, aes(x = Tissue.Diagnosis, y = as.numeric(PCT_INTERGENIC_BASES))) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Fraction Intergenic Bases') + my.theme
multiplot(plotlist = p, cols = 3)
```

@@

### Sample clustering
PCA based clustering of samples
```{r decompse.normalise.data, cache=FALSE, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$E.no.na, scale.=T, center = T)
# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])
plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Tissue.Diagnosis, c('Tissue', 'Diagnosis'), sep = '\\.')
p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
p <- p + theme_bw() + theme(legend.position="top") + facet_grid(Tissue+.~Diagnosis, scales = 'free_y')
p
```

Tree based clustering of samples
```{r decompse.normalise.data.1, cache=FALSE, fig.height=6, fig.width=10, results='asis'}
COVARIATES.tmp = (COVARIATES[,c(FactorCovariates), drop = F])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E.no.na)))
cols = WGCNA::labels2colors(COVARIATES.tmp);
WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

@@

### Sanity check
```{r residual.adj, cache=FALSE, fig.width=12, fig.height=8}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(expr, 
                                              COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
residualSigCovars[["PC_res"]][[2]]$plotData
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(expr))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
PCA of residual data
```{r decompse.normalise.data2.1, cache=FALSE, fig.height=10, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(expr, scale.=T, center = T)
# Plot first 4 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])
plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Tissue.Diagnosis, c('Region','Diagnosis'), sep = '\\.') %>% 
  dplyr::mutate(Region = factor(Region),
                Source = factor(Source))
p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Region~., scales = 'free_y')
p
```
Tree based clustering of residual data
```{r decompse.normalise.data2.2, cache=FALSE, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,FactorCovariates])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
tree = hclust(as.dist(t(expr)))
cols = WGCNA::labels2colors(COVARIATES.tmp);
WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

@@

```{r knitmd, eval=FALSE, cache=FALSE, include=FALSE}
library(synapser)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')
synLogin()
knit2synapse::createAndKnitToFolderEntity(file = "/home/jgockley/TWAS/code/Mayo_RegressDiagnosis_4_TWAS.Rmd",
                                          parentId ="syn20798011",
                                          folderName = 'Mayo')
```
